---
title: "Correlation"
subtitle: "PSYC 640 - Fall 2023"
author: "Dustin Haraden, PhD"
format: 
  revealjs:
    multiplex: true
    scrollable: true
    slide-number: true
    incremental: false
    touch: true
    code-overflow: wrap
    theme: dark
execute: 
  echo: true
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r, include = F}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Last Class

::: nonincremental
-   Two Way ANOVA
    -   Comparing means across multiple groups/levels
:::

------------------------------------------------------------------------

## Looking Ahead

-   R-Workshop! ([Link to Sign up](https://docs.google.com/forms/d/e/1FAIpQLSd2Gjcb88kWHBS7LXNDaLCZ9Sb7aroRceMjrLl5K36epewdtQ/viewform?usp=sf_link))
    -   11/3 & 12/1 from 2-3pm
-   Professor will put together guidelines for the final project and have dates to complete various components
    -   If this point is still here, that means that I haven't gotten to it...

------------------------------------------------------------------------

## Today...

Linear Relationships - Correlation

```{r, results = 'hide', message = F, warning = F}
# for dplyr, ggplot2
library(tidyverse)
#Loading data
library(rio)


#Remove Scientific Notation 
options(scipen=999)
```

# Relationships between variables

## Correlation

Examine the relationship between two continuous variables

Similar to the mean and standard deviation, but it is *between* two variables

Typically displayed as a `scatterplot`

```{r}
# | code-fold: true

set.seed(42)

n <- 200
x <- rnorm(n, mean = 10, sd = 2)
y <- 2 * x + rnorm(n, mean = 0, sd = 2) 
corr_data <- data.frame(x,y)

corr_data %>% 
  ggplot(aes(x,y)) + 
  geom_point() + 
  geom_smooth(method="lm", 
              se = FALSE) + 
  labs(
    x = "Number of Houses", 
    y = "Amount of Candy"
  )


```

------------------------------------------------------------------------

## Statistics and Eugenics

The concept of the correlation is primarily attributed to Sir Frances Galton

-   He was also the founder of the [concept of eugenics](https://www.theguardian.com/commentisfree/2019/oct/03/eugenics-francis-galton-science-ideas)

The correlation coefficient was developed by his student, [Karl Pearson](https://www.britannica.com/biography/Karl-Pearson), and adapted into the ANOVA framework by [Sir Ronald Fisher](https://statmodeling.stat.columbia.edu/2020/08/01/ra-fisher-and-the-science-of-hatred/)

-   Both were prominent advocates for the eugenics movement

------------------------------------------------------------------------

## What do we do with this info?

::: incremental
-   Never use the correlation or the later techniques developed on it? Of course not.

-   Acknowledge this history? Certainly.

-   [Understand how the perspectives](https://medium.com/swlh/is-statistics-racist-59cd4ddb5fa9) of Galton, Fisher, Pearson and others [shaped our practices](http://gppreview.com/2019/12/16/eugenics-ethics-statistical-analysis/)? We must! -- these are not set in stone, [nor are they necessarily the best way](https://www.forbes.com/sites/jerrybowyer/2016/01/06/beer-vs-eugenics-the-good-and-the-bad-uses-of-statistics/?sh=3114a0c82a14) to move forward.
:::

# Next time...

::: incremental
-   ¯\\\_(ツ)\_/¯

-   Correlation & Regression
:::
