---
title: "Correlation"
subtitle: "PSYC 640 - Fall 2023"
author: "Dustin Haraden, PhD"
format: 
  revealjs:
    multiplex: true
    scrollable: true
    slide-number: true
    incremental: false
    touch: true
    code-overflow: wrap
    theme: dark
execute: 
  echo: true
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r, include = F}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Last Class

::: nonincremental
-   Two Way ANOVA
    -   Comparing means across multiple groups/levels
:::

------------------------------------------------------------------------

## Looking Ahead

-   R-Workshop! ([Link to Sign up](https://docs.google.com/forms/d/e/1FAIpQLSd2Gjcb88kWHBS7LXNDaLCZ9Sb7aroRceMjrLl5K36epewdtQ/viewform?usp=sf_link))
    -   11/3 & 12/1 from 2-3pm
-   Final Project Updates:
    -   Introduction & Methods draft due 11/15 (Peer Review)
    -   Data Analysis draft due 11/27 (Peer Review)

------------------------------------------------------------------------

## Today...

Linear Relationships - Correlation

-   Pearson Correlation

-   Spearman's Rank Correlation

-   Missing Data

-   Creating correlation matrices

```{r, results = 'hide', message = F, warning = F}
# for dplyr, ggplot2
library(tidyverse)
#Loading data
library(rio)
# nice tables/plots
library(sjPlot)
library(kableExtra)
library(ggpubr)


#Remove Scientific Notation 
options(scipen=999)
```

# Relationships between variables ([Ch 5.7](https://learningstatisticswithr.com/book/descriptives.html#correl))

## Association - Correlation

Examine the relationship between two continuous variables

Similar to the mean and standard deviation, but it is *between* two variables

Typically displayed as a `scatterplot`

------------------------------------------------------------------------

```{r}
#| code-fold: true

set.seed(42)

n <- 200
x <- rnorm(n, mean = 10, sd = 2)
y <- 2 * x + rnorm(n, mean = 0, sd = 2) 
corr_data <- data.frame(x,y)

corr_data %>% 
  ggplot(aes(x,y)) + 
  geom_point() + 
  geom_smooth(method="lm", 
              se = FALSE) + 
  labs(
    x = "Number of Houses", 
    y = "Amount of Candy"
  )


```

------------------------------------------------------------------------

## Association - Covariance

Before we talk about correlation, we need to take a look at covariance

$$
cov_{xy} = \frac{\sum(x-\bar{x})(y-\bar{y})}{N-1}
$$

::: incremental
-   Covariance can be thought of as the "average cross product" between two variables

-   It captures the *raw/unstandardized* relationship between two variables

-   Covariance matrix is the basis for many statistical analyses
:::

------------------------------------------------------------------------

### Covariance

Let's take a look back at that data before and get the covariance

```{r}
cov(corr_data)
```

::: incremental
-   What does having a covariance of `r round(cov(corr_data)[1,2], 2)` actually mean though?

-   We have to interpret the covariance in terms of the units present (x = \# of houses and y = amount of candy)

    -   The scale is $x*y$ ... what does that even mean?
:::

------------------------------------------------------------------------

## Covariance to Correlation

The Pearson correlation coefficient $r$ addresses this by standardizing the covariance

It is done in the same way that we would create a $z-score$...by dividing by the standard deviation

$$
r_{xy} = \frac{Cov(x,y)}{sd_x sd_y}
$$

------------------------------------------------------------------------

## Correlations

-   Tells us: How much 2 variables are *linearly* related

-   Range: -1 to +1

-   Most common and basic effect size measure

-   Is used to build the regression model

------------------------------------------------------------------------

### Interpreting Correlations ([5.7.5](https://learningstatisticswithr.com/book/descriptives.html#interpretingcorrelations))

| Correlation  |  Strength   | Direction |
|:------------:|:-----------:|:---------:|
| -1.0 to -0.9 | Very Strong | Negative  |
| -0.9 to -0.7 |   Strong    | Negative  |
| -0.7 to -0.4 |  Moderate   | Negative  |
| -0.4 to -0.2 |    Weak     | Negative  |
|  -0.2 to 0   | Negligible  | Negative  |
|   0 to 0.2   | Negligible  | Positive  |
|  0.2 to 0.4  |    Weak     | Positive  |
|  0.4 to 0.7  |  Moderate   | Positive  |
|  0.7 to 0.9  |   Strong    | Positive  |
|  0.9 to 1.0  | Very Strong | Positive  |

# Hypothesis Testing - Correlation

------------------------------------------------------------------------

## Statistical Test - Correlation

We tend to always compare our correlations to the null (0)

::: columns
::: {.column width="50%"}
Hypotheses:

-   $H_0: r_{xy} = 0$

-   $H_1:r_{xy} \neq 0$
:::

::: {.column width="50%"}
Assumptions:

-   Observations are independent

-   Linear Relationship
:::
:::

------------------------------------------------------------------------

## Statistical Test - Correlation

When comparing to 0, we can use the steps similar to a t-test

Calculate the test statistic (df = N - 2):

$$
t = \frac{r}{\sqrt{\frac{1-r^2}{N-2}}}
$$

Then follow the typical steps for a t-test!

## Statistical Test - But not always

It isn't always that easy though...

We were able to use the $t-distribution$ previously because we assumed the null was 0. However, we cannot do that when:

-   The null $\neq$ 0

-   Calculating Confidence Intervals for correlations

-   Comparing two correlations against each other

Will need to transform the $r$ to a $z$ using Fisher's r to z' transformation (beyond this class)

$$
z' = \frac{1}{2}ln\frac{1+r}{1-r}
$$

# Pearson Correlations in R

## Calculating Correlation in R

Now how do we get a correlation value in R?

```{r}
cor(corr_data$x, corr_data$y)

```

That will give us the correlation, but we also want to know how to get our p-value

------------------------------------------------------------------------

### Correlation Test

To get the test of a single pair of variables, we will use the `cor.test()` function:

```{r}
cor.test(x, y, data = corr_data)
```

------------------------------------------------------------------------

## Using real data - NY & NM

So far we have been looking at single variables, but we often care about the relationships between multiple variables in a dataset

```{r}
#| code-fold: true
school <- import("https://raw.githubusercontent.com/dharaden/dharaden.github.io/main/data/NM-NY_CAS.csv") %>% 
  select(Ageyears, Sleep_Hours_Schoolnight, Sleep_Hours_Non_Schoolnight,
         Reaction_time, Score_in_memory_game) %>% 
  janitor::clean_names()

cor(school) %>% 
  kable()
```

# Missing Values

------------------------------------------------------------------------

## Handling Missing - Correlation

-   Listwise Deletion (complete cases)

    ::: incremental
    -   Removes participants completely if they are missing a value being compared
    -   Smaller Sample Sizes
    -   Doesn't bias correlation estimate
    :::

-   Pairwise Deletion

    ::: incremental
    -   Removes participants for that single pair, but leaves information in when there are complete information
    -   Larger Sample Sizes
    -   Could bias estimates if there is a systematic reason things are missing
    :::

------------------------------------------------------------------------

```{r}
cor(school, use = "complete") %>% kable()
```

------------------------------------------------------------------------

```{r}
cor(school, use = "pairwise") %>% kable()
```

# Spearman's Rank Correaltion

------------------------------------------------------------------------

## Shortcomings of Pearson Correlation

::: {style="font-size: 30px"}
Focus on linear relationships - how data fall on a single straight line

::: incremental
-   We assume that with any increase in our X variable, there is an equal amount of increase in Y across the *whole* variable
-   Example: relation between studying/effort and grade
    -   If you put 0 effort you would expect a 0 grade
    -   However, a little bit of effort might be related to a grade of 45
    -   But more effort will need to take place in order to go from 45 - 90 than it does to go from 0 - 45
:::
:::

------------------------------------------------------------------------

## Spearman's Rank Correlation

We need to be able to capture this different (ordinal) "relationship"

-   If student 1 works more hours than student 2, then we can guarantee that student 1 will get a better grade

Instead of using the amount given by the variables ("hours studied"), we rank the variables based on least (rank = 1) to most (rank = 10)

Then we correlate the rankings with one another

# Foundations of Statistics

Who were those white dudes that started this?

------------------------------------------------------------------------

## Statistics and Eugenics

The concept of the correlation is primarily attributed to Sir Frances Galton

-   He was also the founder of the [concept of eugenics](https://www.theguardian.com/commentisfree/2019/oct/03/eugenics-francis-galton-science-ideas)

The correlation coefficient was developed by his student, [Karl Pearson](https://www.britannica.com/biography/Karl-Pearson), and adapted into the ANOVA framework by [Sir Ronald Fisher](https://statmodeling.stat.columbia.edu/2020/08/01/ra-fisher-and-the-science-of-hatred/)

-   Both were prominent advocates for the eugenics movement

------------------------------------------------------------------------

## What do we do with this info?

::: incremental
-   Never use the correlation or the later techniques developed on it? Of course not.

-   Acknowledge this history? Certainly.

-   [Understand how the perspectives](https://medium.com/swlh/is-statistics-racist-59cd4ddb5fa9) of Galton, Fisher, Pearson and others [shaped our practices](http://gppreview.com/2019/12/16/eugenics-ethics-statistical-analysis/)? We must! -- these are not set in stone, [nor are they necessarily the best way](https://www.forbes.com/sites/jerrybowyer/2016/01/06/beer-vs-eugenics-the-good-and-the-bad-uses-of-statistics/?sh=3114a0c82a14) to move forward.
:::

------------------------------------------------------------------------

## Be aware of the assumptions

::: incremental
-   Statistics are often thought of as being absent of bias...they are just numbers

-   Statistical significance was a way to avoid talking about nuance or degree.

-   "Correlation does not imply causation" was a refutation of work demonstrating associations between environment and poverty.

-   Need to be particularly mindful of our goals as scientists and how they can influence the way we interpret the findings
:::

# Fancy Tables

------------------------------------------------------------------------

## Correlation Tables

Before we used the `cor()` function to create a correlation matrix of our variables

***But what is missing?***

```{r}
cor(school, use = "complete") %>% 
  kable()
```

## Correlation Tables - [sjPlot](https://strengejacke.github.io/sjPlot/reference/tab_corr.html)

```{r}
tab_corr(school, na.deletion = "listwise", triangle = "lower")
```

------------------------------------------------------------------------

## Correlation Tables - [sjPlot](https://strengejacke.github.io/sjPlot/reference/tab_corr.html)

So many different cusomizations for this type of plot

Can add titles, indicate what missingness and method

Saves you a TON of time when putting it into a manuscript

# Visualizing Data

------------------------------------------------------------------------

## Visualizing Data

It is always important to visualize our data! Even after getting the correlations and other descriptives

Let's go back to the data that we had in a previous lecture

```{r}
#| code-fold: true
data1 <- import("https://raw.githubusercontent.com/dharaden/dharaden.github.io/main/data/data1.csv") %>% 
  mutate(dataset = "data1")

data2 <- import("https://raw.githubusercontent.com/dharaden/dharaden.github.io/main/data/data2.csv") %>% 
  mutate(dataset = "data2")

data3 <- import("https://raw.githubusercontent.com/dharaden/dharaden.github.io/main/data/data3.csv") %>% 
  mutate(dataset = "data3")
```

And then combine them to make it easier

```{r}
three_data <- bind_rows(data1, data2, data3)
```

------------------------------------------------------------------------

## Descriptive Stats on the 3 datasets

```{r}
three_data %>%
  group_by(dataset) %>% 
  summarize(
    mean_x = mean(x),
    mean_y = mean(y),
    std_x = sd(x), 
    std_y = sd(y), 
    cor_xy = cor(x,y)
  )
```

------------------------------------------------------------------------

## Visualizing Dataset 1

```{r}
#| code-fold: true
data1 %>% 
  ggplot(aes(x, y)) + 
  geom_point() + 
  labs(title = "Dataset 1")
```

------------------------------------------------------------------------

## Visualizing Dataset 1

```{r}
#| code-fold: true
data1 %>% 
  ggplot(aes(x, y)) + 
  geom_point() + 
  labs(title = "Dataset 1") +
  geom_smooth(method = "lm", 
              se = FALSE)
```

------------------------------------------------------------------------

## Visualizing Dataset 1

```{r}
data1 %>% 
  ggscatter("x", "y", 
            add = "reg.line") + 
  stat_cor(label.y = 55) + 
  labs(title = "Dataset 1")
```

------------------------------------------------------------------------

## Visualizing Dataset 2

Let's try it out in R!

# Next time...

-   More Correlation?
-   Maybe Regression? (Y = mX + b)
-   Group Work!
